\chapter{Biocompatibility}
Biocompatibility is essential - if not the most important - aspect to take into conisderation for the specification of the medical device. 
\\
Let's suppose we want ot build a chair. Before building it we must design it in each of its components. We need to figure out how long it is going to be used, where, from whom. The same approach applies to scaffolds and biological devices. A scaffold to be useful should activate specific cellular functions, and each tissue has different chemical and mechanical properties, therefore the scaffold should be designed for the specific region of the body in which it will implanted, taking into consideration the type of cells populating it, the kind of injury, etc.
\\
Back to the chair analogy: for example, which cells will use the device? Cells are the building blocks of biology and can read the information coming from the external enviroment. Given external signaling molecules, specific gene expression is activated to regulate pathways. The enviroment reaches the cell troguh chemical signal, but also through mechanical stimuli. A note on this: scaffold are not \textit{self-sufficient} entities. They are (often) bioactive, but still neeed to to collaborte and make the best out of the enviroment. The ECM composition and the its biocompatibility with the device is as important as the cell-device relationship. 
\\
The biological outcome that will be obtained then depends on a list of parameters:
\begin{description}
\item[Porosity] is extremely important for cell migration. The material should be 3D for the cells, meaning that we should provide a scaffold in which the cells can adhere, grow and migrate into. Take a bottole of water: it's tridimensional for us, but for the cell it's only 2D, as they would adhere and migrate on a flate surface, a plane;
\item[Mechanical properties] are different for each tissue, the scaffold should both resist physical stress but also provide the correct stimulus that the cells need;
\item[Surface] modification is often referred to the functionalization of the scaffold through the addition of proteins (or sequence of aminoacids, we'll talk later about the RGD) to increase adherence;
\item[Antibiotic/antiviral] drugs release system, to control chronic inflammation and possible infections;
\item[Surface topography] may be smooth, rough...
\end{description}
The take home lesson is that the design of a scaffold needs the definition of many parameters that can take many different values. 

\section{Guided tissue regeneration}
In the figure we can see articular cartilage, an intermidiate zone and bone. These three different tissues present differences in vasculartion, cell organization (in cartilage cells live in lacuna, where migration and proliferation are downregulated) and innervation. 
\\
If the damage reaches the bone (which happens most of the times, because the cartilge does not have innervation, while bone does), the scaffold needs to account for different types of tissue. A multicomponent scaffold is designed, something that upregulates angiogenesis in the bone and downregulates it in the cartilage, that has high resistance to mechanical stimulations (cartilage has more water, material reaching water with hydrofilic properties). It should also provide for an enviroment for osteoblasts, lacunae for cartilage cells, space for nerves, vascularity and migration. It also needs to have different degradation times. 
\\
The next step is check the biocompatibility, which will have an impact on cell adhesion, proliferation, differentiation, and also migration; on neoangiogenesis (spefic function) etc. Difference between regeration and repair (no/presence of scar tissue). 

\section{The principles of tissue engineering}
%immagine triangolo
The principles of tissue engineering, the things important to account for can be represented by a pyramid. Designing, choosing the material and in vitro (and then in vivo) testing. \\
Tissue engineering implements a multidisciplinary strategy and the advance in materials' science and biology drastically improve the success of scaffold applicatio. Just consider how much the progressive knowledge in biology incremented the one in biocompatibility. For example, M1 and M2 macrophages were only discovered in 2000. But also the invention of polymers that can change their activity based on the environment (thermo/ph responsive), instructional (functionalization, e.g. with peptides that can control the cells' fate) and that can have specific mechanical strength and function (mechanical signaling). In particular, the materials that can satisfy those requirements are considered 4th generation "biological regenerative biomaterials".
\\
The biomaterials began as inert and stable for their main characteristics, to being able to deliver drugs, to be used to print organs and as cell therapies. The hope for the future is to develop also diagnostic systems and to implant electronic devices, but one thing must always be present: biocompatibility. 

\subsection{Biocompatibility: a definition}
At first, a material is biocompatible when inert, a ghost in the body: "the total absence of interaction between the material
and the tissues". The minimal reaction to the foreign body was preferred, meaning no (or low) inflammatory reaction, no immuno-response. Definition focused on the body reaction to the implant
and what was required was simply the chemical stability by the time.
\\
%immagine valvolve incrostate
People however changed idea based on what they disocvered after seeing what happened in the body after some time after implatation. Heart valve made of titanium and polyester, but the produced failure. Around the plastic layer the body built a very thick scar tissue and the valve could not open anymore. The same valve in another patient induced trombosis. The first valve was treated with anticouagulants. The context of implantation is extremely important! A biological heart valve made of pig, but failed because of calcification, valve was not able to close. All the material induce a biological reaction! No material is completely inert, because our antibodies can recognize anything that is not "self".
\\
The concept of biocompatibility was revised: the ability of a material to perform with an appropriate host response in a specific application” (1987). 
\\
%immagine FBRx
We can take as an example FBRx (cos'è?????): non porous material. Depending on the chemistry of the material we have a specific protein adsorption and subsequently, based on these two, we have an immune response. But if the material is not degradable, the macrophages start to coat the foreign body with a scar tissue. At this point the reaction is finished. If the scaffold degrades eventually we will have an empty bag. Just by changing the porosity the scar tissue is reduced and regeneration is promoted.
\\
Injectable gels (hydrogel) are useful when filling cavities, becomes scaffold and supply for a scaffold for cells to migrate to. 
\\
Take home lesson: scaffold should be tissue-organ dipendent, defined situation, specifically react with the tissue (induce activation of cellular function to heal in a specific enviroment), material degradation. 

\subsubsection{Re-evaluation of the biocompatibility concept}
Some factors led to the redefinition of the concept of biocompatibiliy, as for the scaffold is not enough to simply "exsist":
\begin{itemize}
\item Response to specific materials could vary from one application site to another (tissue-organ dependent) and not just on the material characteristics but it should also defined the situation in which the material is used (pathology, function).
\item An increasing number of applications required that the material specifically react with the tissues rather than be ignored by them. Some applications required that the material degrade over time in the body rather than remain indefinitely.
\item 
\end{itemize}
Most advance concept: biocompatibility is not a material's property but a system's property. It cannot be defined without considering the context.

\section{The complexity of the biocompatible system}
%immagine triangolo con tanta roba
Tissue engineering as a triangle: it seems easy with only three actors. Each one has many possibilities! 
\\
For a scaffold to be biocompatible the following steps must be followed:
\begin{itemize}
\item Determination of native tissue-organ
\item From in vitro to in vivo (from smaller to bigger animals, that depending on the biological problem)
\item At this point we need to define the surgical model and finally we move to clinical trials
\item In addition, for in vitro test now bioreactors are used, beacuse we don't want steady conditions but we want mechanical stimuli (the perfusion provides stree but also washes away cell residues)
\end{itemize}

\subsection{ECM molecules production: effect of the mechanical stimuli}
%Immagine paper
Different sponge-like scaffolds were produced with different properties (like porosity) for cartilage regeneration. 
The quanity of GAG (mainly into the ECM of cartilage) was assesed. In addition, microCT to see where the GAG is and the sponges are completely infilfitraded with it. SL500 had a lot of green (GAG) but only on the outside. What happens if the devices undergoes a mechanical stress similar to the bilogical situations? We get the excat contrary! Just by changing from a static to a perfusion situation. By changing the enviroment we ultimately change integrins and biological performances. In perfusion, the integrins were upregulated! Integrins are transmembrane proteins: portion inside and outside and are important so that scaffolds are able to talk to cells, instructive behaviour. Cells are extremely sensitive. In the very first case the in vitro model  was not well designed. We must perform extensive and well thought tests before moving to the application. 











